Estimating survival in left filtered data

### 2021-12-12 paper.tex
Draft ready

The present project replicates the simulation and bias analyses
presented in Chapter 3 of Izano (2017), embeds the problem in the
non-parametric structural causal approach of Pearl (1995), and applies
the WKM and AWKM estimators to the estimation of colon cancer survival
under exposure/censoring rules in the UAW-GM Cohort.

@@

Figure \ref{fig:survival} presents the true survival curves as well as
the WKM and AWKM survival curves averaged over 250 replications for each
intervention rule and scenario. Qualitatively, the WKM estimator
consistently over-estimated survival whereas the the AWKM survival curve
was much closer to the truth. The bias of the AWKM survival estimator
appeared to be larger in earlier follow-up and smaller for later
follow-up time points. The bias of both the the WKM and the AWKM
estimators appeared largest in Scenario 5.

@@

Table \ref{tab:survival} presents true and estimated average cancer-free
survival times under each intervention rule and scenario. Table
\ref{tab:differences} presents differences in survival time contrasting
rule \(a_1\) to rule \(a_0\). Table \ref{tab:bias} presents estimates of
the bias of the WKM and AWKM estimators for \(\psi\), the difference in
average cancer-free survival time over 20 years of follow-up. These
numeric results were consistent with the qualitative interpretations of
Figure \ref{fig:survival}. The WKM estimator over-estimated the
difference in cancer-free survival time, resulting in bias toward the
null, whereas the AWKM estimator under-estimates the cancer-free
survival, resulting in bias away from the null. In every scenario, the
bias of the WKM-derived contrast was several times larger in magnitude
than that of the AWKM-derived contrast. The bias of both estimators was
greatest for Scenario 5.

@@

In the simulation study, we showed that under several scenarios
compatible with our hypothetical causal structure, the AWKM survival
estimator had small bias compared to the WKM estimator. The bias was
smaller when the cumulative incidence of the outcome was low and at
later follow-up time points. Next, we estimated cancer-free survival in
a real-world context. Using data from the UAW-GM Cohort study, we
followed \(38\,553\) individuals starting from hire to 40 years after
hire for incidence of colon cancer. This duration of follow-up spans the
vast majority of individuals' working lifetimes. As in the simulation,
the UAW-GM data were longitudinal data with baseline covariates,
time-varying covariates, and a survival outcome.

@@

The outcome of interest was colon cancer incidence. There is little past
research linking colon cancer to MWF exposures, but there is some
evidence suggesting that straight MWFs cause colon cancer (Izano et al.
2019). We obtained cancer incidence data via linkage to the Michigan
Cancer Registry, which recorded cancer incidence cases starting on
January 1, 1985. The cohort is comprised of individuals hired between
1938 and 1985. Cancer-free survival to the start of the registry was a
left-filtering process possibly in the presence of the HWSE as
investigated in the simulations. Over the 40 year follow-up period,
vital status was obtained through the Social Security Administration,
the National Death Index, as well as records provided by the UAW.
Baseline confounders included race, sex, plant, and year of hire.
Time-varying confounders included age, cumulative time off, and exposure
to the metalworking fluids not under study. These terms were included in
the estimation of both the treatment and censoring mechanisms, which
were estimated with stratification on every year of follow-up. The
exposure rules of interest were identical to those in the simulation
study: \(a_0\) never exposed and \(a_1\) always exposed while at work.
We assumed that the effect of exposure on health status, mortality,
cancer incidence, etc occurs after a lag of 15 years. Similarly, we
assumed that the effects of cumulative time off occur after a 15 year
lag. Counterfactual survival under rules \(a_0\) and \(a_1\) were
estimated using the WKM and AWKM estimators. Summary statistics for the
full study population and the colon cancer cases are presented in Table
\ref{tab:popchar}.

\begin{table}[h]
\caption{Study population characteristics.}
\label{tab:popchar}
\begin{center}\begin{tabular}{lrlrl}
\toprule
&       \multicolumn{2}{c}{Full cohort}       & \multicolumn{2}{c}{Colon cancer cases}      \\
\midrule
$n$ (person-years) 
            &     38 553 & (1 402 372)       & 267 & (9 635)        \\
Race (\%) \\
\hspace{1.5em} Black   &      7 133 & (18.5)                &  83 & (31.1)          \\
\hspace{1.5em} White   &     31 420 & (81.5)                & 184 & (68.9)          \\
Sex (\%) \\
\hspace{1.5em} Female  &      4 757 & (12.3)                &  49 & (18.4)      \\
\hspace{1.5em} Male    &     33 796 & (87.7)                & 218 & (81.6)      \\
Plant (\%) \\
\hspace{1.5em} Plant 1 &      9 092 & (23.6)                &  66 & (24.7)      \\
\hspace{1.5em} Plant 2 &     17 090 & (44.3)                & 107 & (40.1)      \\
\hspace{1.5em} Plant 3 &     12 371 & (32.1)                &  94 & (35.2)          \\
\midrule
Ever exposed to MWF (\%)
                        &     22193 & (57.6)                & 104 & (39.0)          \\
Year of hire (mean (\textsc{sd}))
                        &     1963 & (11.62)            & 1968 & (8.79)         \\
Age (mean (\textsc{sd}))
                        &     63.38 & (9.08)                & 67.78 & (9.27)        \\
Cumulative time off (mean (\textsc{sd}))
                        &     0.04 & (0.14)               & 0.05 & (0.15)       \\
\bottomrule
\end{tabular}\end{center}
\end{table}

@@

Since we are working with observational data, the evaluation of the
no-interference, causal consistency, ignorability, and positivity
assumptions are critical for causal inference. The stability of our
estimation depends on positivity, which we assessed qualitatively by
examining the distribution of the weights. The no-interference
assumption may be problematized by the fact that there were a finite
number of job types in the factory setting. If one worker operates a
particular metalworking machine, then the the other workers are not able
to operate that machine at that time. Instead, they may be assigned to
assembly tasks, which have lower MWF exposure opportunities. That said,
since these factories were quite large, there may be approximate
independence. The consistency assumption is also problematic. The MWFs
of interest are complex chemical mixtures whose composition changes by
design and by nature. Over the last several decades, the formulation of
MWFs has changed significantly in reaction to performance needs and
toxicity concerns (F. Mirer 2003; Byers 2006). The composition of MWFs
also undergoes unintentional changes because of the nature of their use:
MWFs are often applied in contexts where contamination by other
substances and microbes is possible and chemical changes due to heat and
pressure are likely. Indeed, there are substantial concerns over the
carcinogenicity of the chemical species formed in the MWF mixtures that
were not originally added (Hidajat et al. 2020). Concerns regarding the
consistency assumption may be abrogated in part by adequate adjustment
for secular and factory-level characteristics.

Another key assumption meriting discussion is that of conditional
ignorability. In order to achieve identification, even in the absence of
left filtering, we need to have conditionally ignorable future exposure
status and ignorable future censoring status at each time point given
past data. In occupational cohorts, employment status and health history
are strong predictors of future death (Häfner 1987; Halliday 2014;
Laliotis and Stavropoulou 2018). Logically, major causes of death first
act through employment status before they precipitate death. This
dynamic is actually a key component in the setup for HWSE. We are
therefore relatively confident that conditional ignorability of
censoring due to death is attained given covariate, exposure, and cancer
history. Our confidence in the conditional ignorability of exposure
given history was not as strong. In particular, workers may be assigned
to certain tasks based on their specific skills and knowledge, which may
be associated with structural privileges that confer a lower risk of
deleterious health outcomes. The potential magnitude of this
uncontrolled confounding is likely to be limited, however. The education
level of the workers in the cohort was homogeneous, and all cohort
members were members of the UAW union, which had uniform procedures in
place for equitable access to training, wages, and career advancement
(Harbison 1950; Barnard 2005). The presence of UAW policies support the
assertion that given time since hire, job types (and therefore
exposures) were randomly allocated.

@@

\begin{itemize}
\tightlist
\item
  Foo
\end{itemize}

@@

\begin{itemize}
\tightlist
\item
  FU in GM is much longer than 20 years
\item
  Exclusion restriction on \(H\) so that its parents do not include
  \(A\) is not necessary for identification in full data
\item
  Exclusion restriction on \(D\) so that its parents do not include
  \(H\) not needed if \(H\) is observed and conditioned upon
\item
  Not clear if these were for convenience or what
\item
  What if we increase FU? Lags?
\item
  Greater fineness in exposure definition? Different interventions?
\item
  No way to verify to what extent the simulation informs the real data
  generating system
\item
  Model specification, double robustness
\end{itemize}

@@

>>>>

The present project replicates the simulation and bias analyses
presented in Chapter 3 of Izano (2017), embeds the problem in the
non-parametric structural causal approach of Pearl (1995), and applies
the WKM and AWKM estimators to the estimation of digestive-system
cancer-free survival under exposure/censoring rules in the UAW-GM
Cohort. The \texttt{stremr} package was used to estimate the WKM and
AWKM survival curves in both the simulation and the applied analyses
(Sofrygin, van der Laan, and Neugebauer 2021).

@@

Figure \ref{fig:survival} presents the true survival curves as well as
the WKM and AWKM survival curves averaged over 250 replications for each
intervention rule and scenario. Qualitatively, the WKM estimator
consistently over-estimated survival whereas the the AWKM survival curve
was much closer to the truth. The bias of the AWKM survival estimator
appeared to be larger in earlier follow-up and smaller for later
follow-up time points. The bias of the WKM estimator appeared largest in
Scenario 5. The bias of the AWKM estimator was similar across the five
scenarios

@@

Table \ref{tab:survival} presents true and estimated average cancer-free
survival times under each intervention rule and scenario. Table
\ref{tab:differences} presents differences in survival time contrasting
rule \(a_1\) to rule \(a_0\). Table \ref{tab:bias} presents estimates of
the bias of the WKM and AWKM estimators for \(\psi\), the difference in
average cancer-free survival time over 20 years of follow-up. These
numeric results were consistent with the qualitative interpretations of
Figure \ref{fig:survival}. The WKM estimator over-estimated the
difference in cancer-free survival time, resulting in bias toward the
null, whereas the AWKM estimator under-estimates the cancer-free
survival, resulting in bias away from the null. In every scenario, the
bias of the WKM-derived contrast was two to several times larger in
magnitude than that of the AWKM-derived contrast.

@@

In the simulation study, we showed that under several scenarios
compatible with our hypothetical causal structure, the AWKM survival
estimator had small bias compared to the WKM estimator. The bias was
smaller when the cumulative incidence of the outcome was low and at
later follow-up time points. Next, we estimated cancer-free survival in
a real-world context. Using data from two plants participating in the
UAW-GM Cohort study, we followed \(38\,553\) individuals starting from
hire to 40 years after hire for incidence of digestive system cancers
(colon, rectal, esophageal, or stomach). This duration of follow-up
spans the vast majority of individuals' working lifetimes. As in the
simulation, the UAW-GM data were longitudinal data with baseline
covariates, time-varying covariates, and a survival outcome.

@@

The outcome of interest was digestive system cancer incidence. There is
little past research linking digestive system cancers to MWF exposures,
but there is some evidence suggesting that straight MWFs cause digestive
system cancers (Izano et al. 2019). Cancer incidence was obtained by
linkage to Surveillance, Epidemiology, and End Results (SEER), which
recorded cancer incidence cases starting on January 1, 1973. The cohort
is comprised of individuals hired between 1938 and 1975. Cancer-free
survival to the start of the registry was a left-filtering process
possibly in the presence of the HWSE as investigated in the simulations.
Over the 40 year follow-up period, vital status was obtained through the
Social Security Administration, the National Death Index, as well as
records provided by the UAW. The exposure rules of interest in the
applied analysis were different than those in the simulation study:
\(a_0\) being exposed with 75\% chance in the years exposure was
observed and \(a_1\) being exposed in the years exposure was observed.
Weights were truncated at 1000. Binary exposure is defined to be
exposure above the median level of exposure to straight, soluble, and
synthetic MWFs or not. Counterfactual survival under rules \(a_0\) and
\(a_1\) were estimated using the WKM and AWKM estimators. Baseline
confounders included race, sex, plant, and year of hire. Time-varying
confounders included age, cumulative time off, employment status, and
exposure to the metalworking fluids exposure from the past year. These
terms were included in the estimation of both the treatment and
censoring mechanisms, which were estimated with stratification on every
two years of follow-up. Summary statistics for the full study population
and the digestive system cancer cases are presented in Table
\ref{tab:popchar}.

\begin{table}[h]
\caption{Study population characteristics.}
\label{tab:popchar}
\begin{center}\begin{tabular}{lrlrl}
\toprule
&       \multicolumn{2}{c}{Full cohort}       & \multicolumn{2}{c}{Digestive cancer cases}      \\
\midrule
$n$ (person-years) 
                       &     26 182 & (695 475) & 213 & (6000) \\ 
Race (\%) \\
\hspace{1.5em} Black   &      6 017 & (23.0)  &  66 & (31.0)  \\ 
\hspace{1.5em} White   &     20 165 & (77.0)  & 147 & (69.0)  \\ 
Sex (\%) \\
\hspace{1.5em} Female  &      3 328 & (12.7)  &  15 & (7.0)  \\ 
\hspace{1.5em} Male    &     22 854 & (87.3)  & 198 & (93.0)  \\ 
Plant (\%) \\
\hspace{1.5em} Plant 1 &      9 092 & (34.7)  & 103 & (48.4)  \\ 
\hspace{1.5em} Plant 2 &     17 090 & (65.3)  & 110 & (51.6)  \\ 
\midrule
Ever exposed to MWF (\%)
                                            &      13 240 & (50.6)  &  95 & (44.6)  \\ 
Year of hire (mean (\textsc{sd}))
                                            &            1963 & (12.26) & 1960 & (9.55) \\ 
Age at end of follow-up (mean (\textsc{sd}))
                                            &         55.09 & (12.02) & 63.58 & (9.40) \\ 
Cumulative years off (mean (\textsc{sd}))
                                            &          0.06 & (0.15)  & 0.12 & (0.24)   \\
\bottomrule
\end{tabular}\end{center}
\end{table}

@@

Since we are working with observational data, the evaluation of the
no-interference, causal consistency, ignorability, and overlap
(positivity) assumptions are critical for causal inference. The
stability of our estimation depends on positivity, which we assessed
qualitatively by examining the distribution of the weights. The
no-interference assumption may be problematized by the fact that there
were a finite number of job types in the factory setting. If one worker
operates a particular metalworking machine, then the the other workers
are not able to operate that machine at that time. Instead, they may be
assigned to assembly tasks, which have lower MWF exposure opportunities.
That said, since these factories were quite large, there may be
approximate independence. The consistency assumption is also
problematic. The MWFs of interest are complex chemical mixtures whose
composition changes by design and by nature. Over the last several
decades, the formulation of MWFs has changed significantly in reaction
to performance needs and toxicity concerns (F. Mirer 2003; Byers 2006).
The composition of MWFs also undergoes unintentional changes because of
the nature of their use: MWFs are often applied in contexts where
contamination by other substances and microbes is possible and chemical
changes due to heat and pressure are likely. Indeed, there are
substantial concerns over the carcinogenicity of the chemical species
formed in the MWF mixtures that were not originally added (Hidajat et
al. 2020). Concerns regarding the consistency assumption may be
abrogated in part by adequate adjustment for secular and factory-level
characteristics.

Another key assumption meriting discussion is that of conditional
ignorability. In order to achieve identification, even in the absence of
left filtering, we need to have conditionally ignorable future exposure
status and ignorable future censoring status at each time point given
past data. In occupational cohorts, employment status and health history
are strong predictors of future death (Häfner 1987; Halliday 2014;
Laliotis and Stavropoulou 2018). Logically, major causes of death first
act through employment status before they precipitate death. This
dynamic is actually a key component in the setup for HWSE. We are
therefore relatively confident that conditional ignorability of
censoring due to death is attained given covariate, exposure, and cancer
history. Our confidence in the conditional ignorability of exposure
given history was not as strong. In particular, workers may be assigned
to certain tasks based on their specific skills and knowledge, which may
be associated with structural privileges that confer a lower risk of
deleterious health outcomes. The potential magnitude of this
uncontrolled confounding may have been bounded, however. The education
level of the workers in the cohort was homogeneous, and all cohort
members were members of the UAW union, which had uniform procedures in
place for equitable access to training, wages, and career advancement
(Harbison 1950; Barnard 2005). The presence of UAW policies support the
assertion that given time since hire, job types (and therefore
exposures) were randomly allocated.

@@

Estimated digestive system cancer-free curves under rules \(a_0\) and
\(a_1\) applied to the three MWF types are presented in Figure
\ref{fig:gm_survival}. The survival curves are more or less overlapping
under the two exposure rules. Numeric values for the cancer-free
survival and difference in cancer-free survival are presented in Tables
\ref{tab:gm_survival} and \ref{tab:gm_differences}. The numeric
summaries are consistent with the qualitative interpretation of the
estimated survival curves; for all MWF types, expected survival time
does not differ substantially across rules \(a_0\) and \(a_1\).
Nonetheless, reducing exposure to straight and soluble MWFs does result
in a longer survival time, though the difference is less than a month in
time.

\begin{figure}[h]
\caption{Cancer-free survival over time since hire under interventions for exposure to straight, soluble, and synthetic metalworking fluids. The estimated inverse probability weighted Kaplan-Meier (WKM) survival curve is represented by the solid. The estimated Aalen-filtered inverse probability weighted Kaplan-Meier (AWKM) survival curve is represented by the dashed-line. Salmon color indicates survival and survival estimates under rule $a_0$; cyan color indicates those under rule $a_1$.}
\label{fig:gm_survival}
\begin{center}
\includegraphics{/Users/kevinchen/Documents/2021 Fall/256 - Causal Inference/256-final/reports/resources/gm_results.pdf}
\end{center}
\end{figure}

\begin{table}[H]
\centering
\caption{Estimated cancer-free survival time over 35-year follow-up.} 
\label{tab:gm_survival}
\begin{tabular}{llrrr}
  \toprule
Rule & Estimator & Straight & Soluble & Synthetic \\ 
  \midrule
$a_0$ & WKM & 29.9606 & 29.9605 & 29.9626 \\ 
  $a_0$ & AWKM & 29.9337 & 29.9398 & 29.9443 \\ 
   \midrule
$a_1$ & WKM & 29.9600 & 29.9584 & 29.9648 \\ 
  $a_1$ & AWKM & 29.9325 & 29.9360 & 29.9479 \\ 
   \bottomrule
\end{tabular}
\end{table}
\begin{table}[H]
\centering
\caption{Estimated difference in average cancer-free survival time over 35-year follow-up comparing rule $a_1$ exposed above the median level of exposure with probability 1\% to $a_0$ never exposed.} 
\label{tab:gm_differences}
\begin{tabular}{lrrr}
  \toprule
Estimator & Straight & Soluble & Synthetic \\ 
  \midrule
WKM & -0.0006 & -0.0021 & 0.0022 \\ 
  AWKM & -0.0012 & -0.0039 & 0.0037 \\ 
   \bottomrule
\end{tabular}
\end{table}

Figure \ref{fig:gm_weights} presents the median cumulative weight
applied years since hire with ribbons showing the minimum and maximum
weight. The overlap in the distribution of weights under the two rules
for synthetic MWF exposure suggests that the model for the treatment and
censoring mechanisms were inadequate in distinguishing the units of
analysis by their probability of following a certain exposure rule or of
remaining alive. The distribution of the weights was very skewed.
Without truncation, they would have been in the order of magnitude of
\(10^{20}\) or higher. Distributional summaries of the cumulative
weights without truncation are presented in Table \ref{tab:gm_w}. The
presence of extremely large weights suggests that our observed data were
inadequate for answering the causal question of interest due to
practical violations in overlap (Petersen et al. 2012).

\begin{table}[H]
\centering
\caption{Summary statistics for the cumulative weights used for the WKM estimator at $t = 1$ and $t = 35$ years since hire for each metalworking fluid type and exposure rule.} 
\label{tab:gm_w}
\begin{tabular}{llrrrrrrr}
  \toprule
MWF type & Rule & $t$ & Minimum & Q1 & Median & Mean & Q3 & Maximum \\ 
  \midrule
Straight & $a_0$ &   1 & 1.00 & 1.00 & 1.00 &   1 & 1.00 &   1 \\ 
   & $a_1$ &   1 & 1.00 & 1.00 & 1.00 &   1 & 1.00 &   1 \\ 
  Soluble & $a_0$ &   1 & 1.00 & 1.00 & 1.00 &   1 & 1.00 &   1 \\ 
   & $a_1$ &   1 & 1.00 & 1.00 & 1.00 &   1 & 1.00 &   1 \\ 
  Synthetic & $a_0$ &   1 & 1.00 & 1.00 & 1.00 &   1 & 1.00 &   1 \\ 
   & $a_1$ &   1 & 1.00 & 1.00 & 1.00 &   1 & 1.00 &   1 \\ 
  Straight & $a_0$ &  35 & 0.00 & 2.25 & 4.71 & $9.7 \times 10^{16}$ & 4622.00 & $7.3 \times 10^{20}$ \\ 
   & $a_1$ &  35 & 0.00 & 2.70 & 13.40 & $1.3 \times 10^{18}$ & 22142.45 & $9.8 \times 10^{21}$ \\ 
  Soluble & $a_0$ &  35 & 0.00 & 4.04 & 858.96 & $5.2 \times 10^{15}$ & 74610.19 & $2.3 \times 10^{19}$ \\ 
   & $a_1$ &  35 & 0.00 & 158.56 & 6685.12 & $2.3 \times 10^{17}$ & 1227858.13 & $1.7 \times 10^{21}$ \\ 
  Synthetic & $a_0$ &  35 & 0.00 & 1.02 & 1.25 & $3.7 \times 10^{19}$ & 8.85 & $3.2 \times 10^{23}$ \\ 
   & $a_1$ &  35 & 0.00 & 1.02 & 1.36 & $2.1 \times 10^{21}$ & 22.63 & $1.8 \times 10^{25}$ \\ 
   \bottomrule
\end{tabular}
\end{table}

\begin{figure}
\caption{Median of the cumulative weights for each year since hire with ribbons delimiting the range of values calculated. Salmon color corresponds to the weights used in the estimation of the survival curve under rule $a_0$; cyan corresponds to those for rule $a_1$.}
\label{fig:gm_weights}
\begin{center}
\includegraphics{/Users/kevinchen/Documents/2021 Fall/256 - Causal Inference/256-final/reports/resources/gm_weights.pdf}
\end{center}
\end{figure}

@@

In the simulation study, we showed that the inverse probability of
exposure/censoring weighted Kaplan-Meier estimator resulted in lower
bias in the estimation of average survival time than did the weighted
Kaplan-Meier estimator without the Aalen filter. While the Aalen filter
corrects for non-informative left-filtering in the estimation of the
discrete hazard by excluding person-years not observed under the cancer
registry. These person-years contribute to estimation during the
estimation of exposure and censoring weights. Speculatively, the
combination of the two extensions to the Kaplan-Meier estimator may
result in the partial control of informative left censoring without
adjusting for informative left truncation.

The simulation permitted the evaluation of the estimators in clean,
well-behaved scenarios, but the same luxury was not possible when using
observed data. Several assumptions used in the simulation study were
incompatible with what is known about cancer epidemiology. Firstly,
cumulative incidence in the simulations was close to 15\% (or higher)
over a 20-year follow-up. This is an extraordinarily high risk of
cancer, especially because most workers in the automotive manufacturing
industry start their careers at a fairly young age. Digestive system
cancers are among the most common forms of cancer in the Unite states,
and in 35 years of follow-up, we were still only able to observe a
cumulative incidence of less than 1\%. The simulation scenarios suggest
that the bias of the WKM and AWKM estimators is higher when cumulative
incidence is higher, but when cumulative incidence is low, we rely on
very few observations to index the risk sets.

There are multiple opportunities for future work. The simulation study
can be expanded to investigate subtler exposure-outcome effects. In the
present work, the discrete hazard ratio of cancer incidence given
exposure was rather high at about 1.28. Furthermore, we assumed the
luxury of knowing the true treatment and censoring mechanisms' model
forms, but this is quite unrealistic. In future simulations, the effect
of model mis-specification should be investigated. Alternatively, we
could estimate survival using a doubly robust method. There can also be
several improvements made from the subject-matter expertise
point-of-view. The exposure rules investigated in the applied analysis
were chosen naively. Other dynamic or stochastic exposure interventions
may have been of greater interest to occupational health policy, and
they may have led to improved numeric performance as well. The present
analysis did not lag exposure status, which is standard practice in
cancer epidemiology -- the biological responses to carcinogens do not
occur until several years or decades after exposure. The true causal
effects of MWF exposure may not yet be observable after 35 years of
follow-up.

@@

\leavevmode\hypertarget{ref-Petersen_2012}{}%
Petersen, Maya L, Kristin E Porter, Susan Gruber, Yue Wang, and Mark J
van der Laan. 2012. {``Diagnosing and Responding to Violations in the
Positivity Assumption.''} \emph{Statistical Methods in Medical Research}
21 (1): 31--54. \url{https://doi.org/10.1177/0962280210386207}.

\leavevmode\hypertarget{ref-Sofrygin_2021}{}%
Sofrygin, Oleg, Mark J. van der Laan, and Romain Neugebauer. 2021.
\emph{Stremr: Streamlined Estimation for Static, Dynamic and Stochastic
Treatment Regimes in Longitudinal Data}.
\url{https://github.com/osofr/stremr}.

### 2021-12-12 reports/01-introduction.Rmd
Draft ready

The present project replicates the simulation and bias analyses presented in Chapter 3 of @Izano_2017_thesis, embeds the problem in the non-parametric structural causal approach of @Pearl_1995, and applies the WKM and AWKM estimators to the estimation of colon cancer survival under exposure/censoring rules in the UAW-GM Cohort.

>>>>

The present project replicates the simulation and bias analyses presented in Chapter 3 of @Izano_2017_thesis, embeds the problem in the non-parametric structural causal approach of @Pearl_1995, and applies the WKM and AWKM estimators to the estimation of digestive-system cancer-free survival under exposure/censoring rules in the UAW-GM Cohort. The `stremr` package was used to estimate the WKM and AWKM survival curves in both the simulation and the applied analyses [@Sofrygin_2021].

### 2021-12-12 reports/03-results.Rmd
Draft ready

Table \ref{tab:survival} presents true and estimated average cancer-free survival times under each intervention rule and scenario. Table \ref{tab:differences} presents differences in survival time contrasting rule $a_1$ to rule $a_0$. Table \ref{tab:bias} presents estimates of the bias of the WKM and AWKM estimators for $\psi$, the difference in average cancer-free survival time over 20 years of follow-up. These numeric results were consistent with the qualitative interpretations of Figure \ref{fig:survival}. The WKM estimator over-estimated the difference in cancer-free survival time, resulting in bias toward the null, whereas the AWKM estimator under-estimates the cancer-free survival, resulting in bias away from the null. In every scenario, the bias of the WKM-derived contrast was several times larger in magnitude than that of the AWKM-derived contrast. The bias of both estimators was greatest for Scenario 5.

>>>>

Table \ref{tab:survival} presents true and estimated average cancer-free survival times under each intervention rule and scenario. Table \ref{tab:differences} presents differences in survival time contrasting rule $a_1$ to rule $a_0$. Table \ref{tab:bias} presents estimates of the bias of the WKM and AWKM estimators for $\psi$, the difference in average cancer-free survival time over 20 years of follow-up. These numeric results were consistent with the qualitative interpretations of Figure \ref{fig:survival}. The WKM estimator over-estimated the difference in cancer-free survival time, resulting in bias toward the null, whereas the AWKM estimator under-estimates the cancer-free survival, resulting in bias away from the null. In every scenario, the bias of the WKM-derived contrast was two to several times larger in magnitude than that of the AWKM-derived contrast.
